---
title: "Region 5 results"
author: "Jess Kunke"
date: "Nov 2025"
format: pdf
  # pdf:
  #   geometry:
  #     - top=20mm
  #     - left=20mm
  #     - right=20mm
  #     - bottom=30mm
toc: true
execute: 
  echo: false
  message: false
  warning: false
---

```{r setup, include = FALSE}
library(tidyverse)
library(knitr)
library(gridExtra)
library(tictoc)
```

```{r}
# load("pwdists/input/preds_obs_pwdist_input_data_Region5.RData")
# load("pwdists/input/preds_obs_fish_for_estimation_Region5.RData")
# load("Region5_results/Region5_est_pred_results.RData")
# load("Region5_results/nobs_by_species.rda")
# 
# # update Nov 14 2025: remove 24 species that are identified only at the genus level
# results_by_species = results_by_species[!str_detect(tolower(names(results_by_species)), "genus")]
# results_by_species = results_by_species[!str_detect(tolower(names(results_by_species)), "subspecies")]
# 
# q = nrow(preds)
# n = n_distinct(fish_region5$COMID)
# # update Nov 14 2025: remove 24 species that are identified only at the genus level
# nobs_by_species = nobs_by_species[!str_detect(tolower(nobs_by_species$Common_Name), "genus"),]
# nobs_by_species = nobs_by_species[!str_detect(tolower(nobs_by_species$Common_Name), "subspecies"),]
# n_species = nrow(nobs_by_species)
# 
# # col 1 = COMID, cols 2 through end are the 309 different species observed in Region 5
# density_by_site_obs = matrix(nrow = n, ncol = n_species+1)
# density_by_site_obs[,1] = preds$COMID[1:n]
# 
# density_by_site_pred = matrix(nrow = n, ncol = n_species+1)
# density_by_site_pred[,1] = preds$COMID[1:n]
# 
# total_counts = data.frame(
#   # species name
#   species = nobs_by_species$Common_Name,
#   # number of COMIDs on which this species was observed
#   nCOMIDs = nobs_by_species$nCOMIDs,
#   # proportion of COMIDs on which any fish was observed on which
#   # *THIS* species was observed
#   p_COMIDs_with_obs = nobs_by_species$nCOMIDs/n,
#   # total observed count
#   obs_total_count = nobs_by_species$TotalCount,
#   # total predicted count on COMIDs where *any fish* were observed
#   pred_total_count_wherefishobs = NA,
#   # total predicted count on COMIDs where *this species* was observed
#   pred_total_count_wherethisspecobs = NA,
#   # total predicted Region 5 population size
#   # (sum of 10*DensityPer100m*LENGTHKM over COMIDs with nonnegative predicted densities)
#   regional_estimate = NA,
#   # proportion of predicted densities that are <= -1
#   # total_counts$p_neg[s] = sum(DensityPer100m <= -1)/q
#   p_neg = NA,
#   # proportion of predicted densities *on COMIDs with observations of any
#   # fish species* that are <= -1
#   # sum(DensityPer100m[1:n] <= -1)/n
#   p_neg_obs = NA,
#   # proportion of predicted densities *on COMIDs with observations of any
#   # fish species* for which (1) predicted density <= -1 *AND* (2) this species
#   # was not observed
#   # sum(DensityPer100m[1:n][obs_density==0] < -1)/n
#   p_neg_obs0 = NA,
#   # median of the negative predicted densities
#   # median(DensityPer100m[1:n][DensityPer100m[1:n] < 0])
#   median_neg = NA,
#   # median of the positive predicted densities
#   # median(DensityPer100m[1:n][DensityPer100m[1:n] > 0])
#   median_pos = NA,
#   # correlation between predicted and observed density on COMIDs in which
#   # any fish species was observed
#   # cor(DensityPer100m[1:n], obs_density)
#   cor_obs_pred = NA,
#   # correlation between predicted and observed density on COMIDs in which
#   # *THIS* fish species was observed
#   # cor(DensityPer100m[1:n][obs_density > 0], obs_density[obs_density > 0])
#   cor_obs_pred_nz = NA,
#   # estimation runtime
#   estimation_time = NA,
#   estimation_units = NA,
#   # prediction runtime
#   prediction_time = NA,
#   prediction_units = NA
# )
# 
# params = data.frame(
#   beta_1 = NA,
#   beta_2 = NA,
#   beta_3 = NA,
#   beta_4 = NA,
#   beta_5 = NA,
#   beta_6 = NA,
#   beta_7 = NA,
#   beta_8 = NA,
#   beta_9 = NA,
#   beta_10 = NA,
#   theta_1 = NA,
#   theta_2 = NA,
#   theta_3 = NA
# )
# 
# tic("diagnostics loop"); for(s in 1:nrow(total_counts)){ #for(s in 1:10){
#   cat(paste0(s, " "))
#   species = total_counts$species[s]
# 
#   if(!is.na(results_by_species[[s]][1])){
# 
#     DensityPer100m = results_by_species[[s]]$preds_supp$DensityPer100m_pred
#     CountReachTotal = results_by_species[[s]]$preds_supp$CountReachTotal_pred
#     obs_density = results_by_species[[s]]$preds_supp$obs_density
# 
#     # predicted total count
#     total_counts$pred_total_count_wherefishobs[s] = sum(CountReachTotal[1:n][CountReachTotal[1:n] >=0])
#     total_counts$pred_total_count_wherethisspecobs[s] = sum(CountReachTotal[1:n][CountReachTotal[1:n] >=0 & obs_density[1:n] > 0])
#     total_counts$regional_estimate[s] = sum(CountReachTotal[CountReachTotal >=0])
#     total_counts$p_neg[s] = sum(DensityPer100m <= -1)/q
#     total_counts$p_neg_obs[s] = sum(DensityPer100m[1:n] <= -1)/n
#     total_counts$p_neg_obs0[s] = sum(DensityPer100m[1:n][obs_density[1:n]==0] < -1)/n
#     total_counts$median_neg[s] = median(DensityPer100m[1:n][DensityPer100m[1:n] < 0])
#     total_counts$median_pos[s] = median(DensityPer100m[1:n][DensityPer100m[1:n] > 0])
#     total_counts$cor_obs_pred[s] = cor(DensityPer100m[1:n], obs_density[1:n])
#     total_counts$cor_obs_pred_nz[s] = cor(DensityPer100m[1:n][obs_density[1:n] > 0],
#                                           obs_density[1:n][obs_density[1:n] > 0])
#     # ensure all estimation and predictions times have the same units
#     units(results_by_species[[s]]$estimation_time) = "secs"
#     units(results_by_species[[s]]$prediction_time) = "secs"
#     total_counts$estimation_time[s] = results_by_species[[s]]$estimation_time
#     total_counts$prediction_time[s] = results_by_species[[s]]$prediction_time
#     total_counts$prediction_units[s] = "secs"
#     total_counts$estimation_units[s] = "secs"
#     params[s,] = c(results_by_species[[s]]$estimation$Beta, results_by_species[[s]]$estimation$Theta)
# 
#     density_by_site_pred[, s+1] = DensityPer100m[1:n]
#     density_by_site_obs[, s+1] = obs_density[1:n]
#   }
# }; cat("\n"); toc()
# # diagnostics loop: 55.643 sec elapsed
# 
# diagnostics = cbind(total_counts, params) %>%
#   dplyr::rename(
#     Intercept = beta_1,
#     Elevation = beta_2,
#     Water_Area = beta_3,
#     Ann_Runoff = beta_4,
#     Baseflow = beta_5,
#     Ann_Temp = beta_6,
#     Development = beta_7,
#     Agriculture = beta_8,
#     Hydro_Alter = beta_9,
#     Fldplain_Dis = beta_10,
#     SigmaSq = theta_1,
#     TauSq = theta_2,
#     Phi = theta_3
#   )
# 
# cor_density_by_site = rep(NA, n)
# for(i in 1:n){
#   cor_density_by_site[i] = cor(density_by_site_obs[i,2:(n_species+1)],
#                                density_by_site_pred[i,2:(n_species+1)])
# }
# 
# save(diagnostics, cor_density_by_site,
#      file = "Region5_diagnostics.RData")
```

```{r}
load("Region5_diagnostics.RData.RData")
n_species = nrow(diagnostics)
plot_dir = "Region5_results/"
```

:::{.callout-note collapse="true"}
## More about the data sources

All of this data is available online; we'll provide sources in the paper.

- [elevation and basin area](https://www.sciencebase.gov/catalog/item/57976a0ce4b021cadec97890)
- [development and agriculture](https://www.sciencebase.gov/catalog/item/5d66b3b6e4b0c4f70cefb11d)
- [annual runoff](https://www.sciencebase.gov/catalog/item/578f8ad8e4b0ad6235cf6e43)
- [baseflow](https://www.sciencebase.gov/catalog/item/5669a8e3e4b08895842a1d4f)
- [temperature](https://www.sciencebase.gov/catalog/item/566ef828e4b09cfe53ca76f7)
- [hydrological alteration](https://www.nature.com/articles/s41597-022-01566-1)
- [floodplain integrity](https://www.nature.com/articles/s43247-023-00877-4)

Citation for most of the data: Wieczorek, M.E., Jackson, S.E., and Schwarz, G.E., 2018, Select Attributes for NHDPlus Version 2.1 Reach Catchments and Modified Network Routed Upstream Watersheds for the Conterminous United States (ver. 4.0, August 2023): U.S. Geological Survey data release, https://doi.org/10.5066/F7765D7V.
:::


## Runtime

```{r}
#| label: fig-runtimes
#| fig-cap: "Computational expense for estimating and predicting."
#| fig-subcap: 
#|   - "Runtime for estimating model parameters"
#|   - "Runtime for predicting fish densities across Region 5"
#| layout: "[45,-10,45]"

ggplot(diagnostics, aes(x = estimation_time)) + 
  geom_histogram(bins = 20) +
  xlab("Estimation time (seconds)") +
  ylab("Count") +
  theme_bw()

ggsave(filename = paste0(plot_dir, "est_time.png"),
       width = 3, height = 3, units = "in")

ggplot(diagnostics, aes(x = prediction_time)) + 
  geom_histogram(bins = 20) +
  xlab("Prediction time (seconds)") +
  ylab("Count") +
  theme_bw()

ggsave(filename = paste0(plot_dir, "pred_time.png"),
       width = 3, height = 3, units = "in")

# est_time = ggplot(diagnostics, aes(x = estimation_time)) + 
#   geom_histogram(bins = 20) +
#   xlab("Estimation time (seconds)") +
#   theme_bw()
# 
# pred_time = ggplot(diagnostics, aes(x = prediction_time)) + 
#   geom_histogram(bins = 20) +
#   xlab("Prediction time (seconds)") +
#   theme_bw()
# 
# both = grid.arrange(est_time, pred_time, nrow = 1)
# 
# ggsave(both, filename = paste0(plot_dir, "est_pred_times.png"),
#        width = 8, height = 3, units = "in")
```

@fig-runtimes shows the distribution of estimation and prediction runtimes across species. On average, estimation and prediction for a given species took `r round(mean(diagnostics$estimation_time), 2)` seconds and `r round(mean(diagnostics$prediction_time), 2)` seconds, respectively. Total estimation and prediction time across all 309 species was `r round((sum(diagnostics$prediction_time) + sum(diagnostics$estimation_time))/3600, 2)` hours.

## Number of observations per species

@fig-npCOMIDs shows the number and proportion of COMIDs with any fish observations that have observations of each species.

The number of COMIDs with observations of each species ranges from `r min(diagnostics$nCOMIDs)` to `r max(diagnostics$nCOMIDs)` (mean `r round(mean(diagnostics$nCOMIDs), 1)`; quartiles `r quantile(diagnostics$nCOMIDs, 0.25)`, `r quantile(diagnostics$nCOMIDs, 0.5)`, `r quantile(diagnostics$nCOMIDs, 0.75)`).

The proportion of COMIDs with observations of at least one species that have observations of a given species ranges from `r format(round(min(diagnostics$p_COMIDs_with_obs), 4), scientific=FALSE)` to `r format(round(max(diagnostics$p_COMIDs_with_obs), 4), scientific=FALSE)` (mean `r format(round(mean(diagnostics$p_COMIDs_with_obs), 4), scientific=FALSE)`; quartiles `r format(round(quantile(diagnostics$p_COMIDs_with_obs, 0.25), 4), scientific=FALSE)`, `r format(round(quantile(diagnostics$p_COMIDs_with_obs, 0.5), 4), scientific=FALSE)`, `r format(round(quantile(diagnostics$p_COMIDs_with_obs, 0.75), 4), scientific=FALSE)`).

```{r}
#| label: fig-npCOMIDs
#| fig-cap: "Number and proportion of COMIDs with observations of each species."
#| fig-subcap: 
#|   - "Number of COMIDs with observations of each species"
#|   - "Proportion of COMIDs with any fish observations that have observations of each species"
#| layout: "[45,-10,45]"

ggplot(diagnostics, aes(x = nCOMIDs)) + 
  geom_histogram(bins = 20) +
  xlab("Number of COMIDs on which this species was observed") +
  theme_bw()

ggplot(diagnostics, aes(x = p_COMIDs_with_obs)) +
  geom_histogram(bins = 20) +
  xlab("Proportion") +
  theme_bw()
```


## Diagnostics

### Total observed versus predicted fish counts

@fig-total-counts-histograms shows the distribution of observed and predicted counts across species, and @fig-total-counts-scatterplots shows scatterplots of observed versus predicted counts. I show three possible predicted counts: 

a. totals across only Region 5 COMIDs at which the species in question was observed,
b. totals across only Region 5 COMIDs at which any fish were observed, and
c. regional estimates (totals across all Region 5 COMIDs).

The ratio of predicted to observed counts range from `r round(min(diagnostics$pred_total_count_wherethisspecobs/diagnostics$obs_total_count), 2)` to `r round(max(diagnostics$pred_total_count_wherethisspecobs/diagnostics$obs_total_count), 2)` at COMIDs where the species in questions was observed, from `r round(min(diagnostics$pred_total_count_wherefishobs/diagnostics$obs_total_count), 2)` to `r round(max(diagnostics$pred_total_count_wherefishobs/diagnostics$obs_total_count), 2)` at COMIDs where any fish were observed, and from `r round(min(diagnostics$regional_estimate/diagnostics$obs_total_count), 2)` to `r round(max(diagnostics$regional_estimate/diagnostics$obs_total_count), 2)` using the regional estimates as the predicted counts.

```{r}
#| label: fig-total-counts-histograms
#| fig-cap: "Observed and predicted total counts across Region 5."
#| fig-subcap: 
#|   - "Distribution of observed counts"
#|   - "Distribution of predicted counts (over all Region 5 COMIDs at which the species in question was observed)"
#|   - "Distribution of predicted counts (over all Region 5 COMIDs at which any fish were observed)"
#|   - "Distribution of regional estimates (over all Region 5 COMIDs)"
#| layout: "[[45,-10,45],[45,-10,45]]"

ggplot(diagnostics, aes(x = obs_total_count/10^6)) + 
  geom_histogram(bins = 20) +
  xlab("Total observed count (millions of fish)") +
  theme_bw()

ggplot(diagnostics, aes(x = pred_total_count_wherethisspecobs/10^6)) + 
  geom_histogram(bins = 20) +
  xlab("Total predicted count at COMIDs with obs of this species (millions of fish)") +
  theme_bw()

ggplot(diagnostics, aes(x = pred_total_count_wherefishobs/10^6)) + 
  geom_histogram(bins = 20) +
  xlab("Total predicted count at COMIDs with any obs (millions of fish)") +
  theme_bw()

ggplot(diagnostics, aes(x = regional_estimate/10^6)) + 
  geom_histogram(bins = 20) +
  xlab("Total predicted count (millions of fish)") +
  theme_bw()
```


```{r}
#| label: fig-total-counts-scatterplots
#| fig-cap: "Observed vs predicted total counts across Region 5."
#| fig-subcap: 
#|   - "Observed counts vs predicted counts where the species in question was observed"
#|   - "Observed counts vs predicted counts where any fish were observed"
#|   - "Observed counts vs predicted regional estimates"
#| layout: "[[45,-10,45],[45,-10,45]]"
ggplot(diagnostics, aes(x = obs_total_count/10^6, y = pred_total_count_wherethisspecobs/10^6)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  xlab("Observed total count (millions of fish)") +
  ylab("Predicted total count (millions of fish)") +
  theme_bw()
  
ggsave(filename = paste0(plot_dir, "obs_v_pred_wherethisspec.png"),
  width = 3, height = 3, units = "in")

ggplot(diagnostics, aes(x = obs_total_count/10^6, y = pred_total_count_wherefishobs/10^6)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  xlab("Observed total count (millions of fish)") +
  ylab("Predicted total count (millions of fish)") +
  theme_bw()
  
ggsave(filename = paste0(plot_dir, "obs_v_pred_wherefishobs.png"),
  width = 3, height = 3, units = "in")

ggplot(diagnostics, aes(x = obs_total_count/10^6, y = regional_estimate/10^6)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  xlab("Observed total count (millions of fish)") +
  ylab("Predicted total count (millions of fish)") +
  theme_bw()
  
ggsave(filename = paste0(plot_dir, "obs_v_pred.png"),
  width = 3, height = 3, units = "in")
```

### Negative predicted densities

@fig-neg-dens-prop shows the distribution of negative predicted densities. I updated these results to consider densities less than or equal to -1. Of the `r n_species` species, `r sum(diagnostics$p_neg == 0)` (`r round(100*mean(diagnostics$p_neg == 0))`%) have no negative predicted densities. The mean and max proportion of negative predicted densities by species are `r round(mean(diagnostics$p_neg_obs), 4)` and `r round(max(diagnostics$p_neg_obs), 4)`, respectively.

```{r}
#| label: fig-neg-dens-prop
#| fig-cap: "Proportions of densities that are negative."
#| fig-subcap: 
#|   - "Distribution of the proportion of predicted densities that are less than or equal to -1"
#|   - "Same as (a), except the denominator is the number of COMIDs with observations of any fish species"
#|   - "Distribution of the proportion in which the numerator is the number of COMIDs in which this species was not observed and its predicted density is less than or equal to -1, and the denominator is the number of COMIDs with observations of any fish species"
#| layout: "[[45,-10,45],[45,-10,45]]"

ggplot(diagnostics, aes(x = p_neg)) + 
  geom_histogram(bins = 20) +
  xlab("Proportion") +
  theme_bw()

ggplot(diagnostics, aes(x = p_neg_obs)) + 
  geom_histogram(bins = 20) +
  xlab("Proportion") +
  theme_bw()

ggplot(diagnostics, aes(x = p_neg_obs0)) + 
  geom_histogram(bins = 20) +
  xlab("Proportion") +
  theme_bw()
```

```{r}
#| eval: false

# sum(DensityPer100m <= -1)/q
summary(diagnostics$p_neg)
sum(diagnostics$p_neg == 0); mean(diagnostics$p_neg == 0)

# sum(DensityPer100m[1:n] <= -1)/n
summary(diagnostics$p_neg_obs)
sum(diagnostics$p_neg_obs == 0); mean(diagnostics$p_neg_obs == 0)

# sum(DensityPer100m[1:n][obs$Count==0] < -1)/n
summary(diagnostics$p_neg_obs0)
sum(diagnostics$p_neg_obs0 == 0); mean(diagnostics$p_neg_obs0 == 0)
```

I also computed the median positive and negative densities by species. @tbl-median-densities summarizes the median positive and negative densities by species, and @fig-neg-dens-medians shows distributions and scatterplots of the median positive and negative densities by species.

```{r}
#| label: tbl-median-densities
#| tbl-cap: "Summary of median positive and negative densities by species."

tbl_medians = rbind(summary(diagnostics$median_neg),
                    summary(diagnostics$median_pos))

rownames(tbl_medians) = c("Negative", "Positive")
colnames(tbl_medians) = names(summary(diagnostics$median_neg))

kable(tbl_medians, digits = 4)
```

```{r}
#| label: fig-neg-dens-medians
#| fig-cap: "Median positive and negative densities by species."
#| fig-subcap: 
#|   - "Distribution of the median negative density"
#|   - "Distribution of the median positive density"
#|   - "Distribution of the ratio of median negative density to median positive density"
#|   - "Median negative density vs median positive density"
#| layout: "[[45,-10,45],[45,-10,45]]"

ggplot(diagnostics, aes(x = median_neg)) + 
  geom_histogram(bins = 20) +
  xlab("Median negative density") +
  theme_bw()

ggplot(diagnostics, aes(x = median_pos)) + 
  geom_histogram(bins = 20) +
  xlab("Median positive density") +
  theme_bw()

ggplot(diagnostics, aes(x = median_neg/median_pos)) + 
  geom_histogram(bins = 20) +
  labs(x = "Ratio", y = "Count") +
  theme_bw()
  
ggsave(filename = paste0(plot_dir, "median_neg_pos_density_ratio.png"),
  width = 3, height = 3, units = "in")

ggplot(diagnostics, aes(y = median_neg, x = median_pos)) + 
  geom_point() +
  ylab("Median negative density") +
  xlab("Median positive density") +
  theme_bw()
  
ggsave(filename = paste0(plot_dir, "median_neg_vs_pos_density.png"),
  width = 3, height = 3, units = "in")
```


### Correlations by species

@tbl-cor-by-species and @fig-cor-by-species show the correlations between predicted and observed densities across all sites for each species. These correlations use the predicted densities without any modification, i.e. with negative predicted densities included.

```{r}
#| label: tbl-cor-by-species
#| tbl-cap: "Correlations by species."

tbl_cor_by_spec = data.frame(matrix(summary(diagnostics$cor_obs_pred), nrow=1))
names(tbl_cor_by_spec) = names(summary(diagnostics$cor_obs_pred))

sd(tbl_cor_by_spec)

kable(tbl_cor_by_spec, digits = 3)
```

```{r}
#| label: fig-cor-by-species
#| fig-cap: "Correlations between observed and predicted fish densities by species across Region 5."
#| fig-subcap: 
#|   - "Correlations by species across all of Region 5"
#|   - "Correlations by species only for COMIDs at which this species was observed"
#| layout: "[45,-10,45]"

ggplot(diagnostics, aes(x = cor_obs_pred)) + 
  geom_histogram(aes(y = after_stat(count/sum(count))), bins = 20) +
  labs(x = "Correlation", y = "Proportion of species") +
  theme_bw()
  
ggsave(filename = paste0(plot_dir, "cor_by_species.png"),
  width = 3, height = 3, units = "in")

ggplot(diagnostics, aes(x = cor_obs_pred_nz)) + 
  geom_histogram(bins = 20)
```

### Correlations by site

@tbl-cor-by-site and @fig-cor-by-site show the correlations between predicted and observed densities for all species at each site.

```{r}
#| label: tbl-cor-by-site
#| tbl-cap: "Correlations by sites."

tbl_cor_by_site = data.frame(matrix(summary(cor_density_by_site), nrow=1))
names(tbl_cor_by_site) = names(summary(cor_density_by_site))

sd(cor_density_by_site)

kable(tbl_cor_by_site, digits = 3)
```

```{r}
#| label: fig-cor-by-site
#| fig-cap: "Correlations between observed and predicted fish densities by site across all species."
#| fig-width: 4
#| fig-height: 2

ggplot(data.frame(cor_density_by_site), aes(cor_density_by_site)) + 
  geom_histogram(aes(y = after_stat(count/sum(count))), bins = 20) +
  labs(x = "Correlation", y = "Proportion of sites") +
  theme_bw()
  
ggsave(filename = paste0(plot_dir, "cor_by_site.png"),
  width = 3, height = 3, units = "in")
```


